---
title: "Plastic Waste Dashboard"
format: 
  dashboard:
    orientation: columns
    theme: flatly
embed-resources: true
logo: plastic-logo.svg
logo-alt: "Recycling Logo with caption"
nav-buttons:
      - icon: github
        href: https://github.com/FaranIgani
        aria-label: GitHub
      - icon: linkedin
        href: https://linkedin.com/in/faran-igani
        aria-label: LinkedIn
---

```{r}
#| label: load-packages
library(tidyverse)
library(gt)
library(rvest)
library(ggplot2)
library(dplyr)
library(httr2)
library(jsonlite)
library(httr)
library(glue)
library(tidyjson)
library(ggridges)
library(scales)
library(stringdist)
library(tidytext)
library(leaflet)
library(maps)
library(plotly)
library(gganimate)
library(gifski)
```

```{r}
#| label: Load the Data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

```{r}
#| label: data-from-api
#

base_url <- "https://restcountries.com/v4/"

restcountries <- GET(glue("{base_url}all?fields=name,continents,region,subregion,unMember,gini,density,geolocation"))

restcountries_tidy <- restcountries %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  as_tibble()

```

```{r}
#| label: merge-plastics

library(stringi)

normalize_name <- function(x) {
  x %>% 
    stri_trans_general("Latin-ASCII") %>% 
    str_to_lower() %>% 
    str_remove_all("[^a-z ]") %>% 
    str_trim() %>% 
    str_squish()
}

plastics_clean <- plastics %>% 
  filter(country != "EMPTY") %>% 
  mutate(
    country = case_when(
      country == "Korea" ~ "South Korea",
      country == "Taiwan_ Republic of China (ROC)" ~ "Taiwan",
      country == "United Kingdom of Great Britain & Northern Ireland" ~ 
        "United Kingdom",
      TRUE ~ country
    )
  ) %>%
  mutate(country_key = normalize_name(country))

restcountries_clean <- restcountries_tidy %>% 
  select(name.common, name.official, region, subregion,
         density, unMember, geolocation.latitude, 
         geolocation.longitude) %>% 
  mutate(common_key = recode(
    normalize_name(name.common),
    "ivory coast" = "cote divoire"
    ),
         official_key = normalize_name(name.official))
  

# Different name types, try both

joined <- plastics_clean %>%
  left_join(
    restcountries_clean %>% select(-official_key),
    by = c("country_key" = "common_key")
  )

matched <- joined %>% filter(!is.na(region))
unmatched <- joined %>% filter(is.na(region)) %>%
  select(names(plastics_clean))

refilled <- unmatched %>%
  left_join(
    restcountries_clean %>% select(-common_key),
    by = c("country_key" = "official_key")
  )

joined1 <- bind_rows(matched, refilled)
```

```{r}
#| label: merge-plastics-cont
## Fuzzy matching for parent companies using stringdist

noise_words <- c(
  "\\bthe\\b",
  "\\binc\\.?\\b",
  "\\bco\\.?\\b",
  "\\bltd\\.?\\b",
  "\\bcorp\\.?\\b",
  "\\bcompany\\b",
  "\\bcompanies\\b",
  "\\bgroup\\b",       # "Nestlé Group" -> "nestlé"
  "\\bholdings\\.?\\b",
  "\\bllc\\.?\\b",
  "\\bplc\\.?\\b",
  "\\bgmbh\\b",        # common German suffix
  "\\bsa\\b",          # common in Latin America/Europe
  "\\bsrl\\b"
)

normalize_company <- function(x) {
  x %>%
    str_to_lower() %>%
    str_remove_all(paste(noise_words, collapse = "|")) %>%
    str_remove_all("[^a-z0-9 ]") %>%
    str_squish()
}

resolve_typos <- function(names, threshold = 0.1) {
  dist_matrix <- stringdistmatrix(names, names, method = "jw")
  diag(dist_matrix) <- Inf
  
  canonical <- names
  
  for (i in seq_along(names)) {
    best_dist <- min(dist_matrix[i, ])
    if (best_dist < threshold) {
      best_match <- which.min(dist_matrix[i, ])
      canonical[i] <- min(names[i], names[best_match])
    }
  }
  
  tibble(normalized = names, canonical = str_to_title(canonical))
}

companies_raw <- unique(joined1$parent_company)

name_map <- tibble(original = companies_raw) %>%
  mutate(normalized = normalize_company(original)) %>%
  left_join(
    resolve_typos(unique(.$normalized)),  # fuzzy match on normalized names only
    by = "normalized"
  )

canonical_labels <- name_map %>%
  count(canonical, original) %>%          # count how often each original appears
  slice_max(n, by = canonical, n = 1, with_ties = FALSE) %>% # pick the most frequent per canonical
  select(canonical, display_name = original)

joined <- joined1 %>%
  left_join(name_map %>% select(original, canonical),
            by = c("parent_company" = "original")) %>%
  left_join(canonical_labels, by = "canonical") %>% 
  mutate(parent_company = display_name) %>% 
  select(-display_name, -canonical)
```

```{r}
#| label: API, World Bank API

base_url <- "https://api.worldbank.org/v2/country/all/indicator/"

pop_code <- "SP.POP.TOTL"
gdp_code <- "NY.GDP.MKTP.CD"

pop_url <- str_c(base_url, pop_code, "?format=json&per_page=20000")
gdp_url <- str_c(base_url, gdp_code, "?format=json&per_page=20000")

pop_raw <- GET(pop_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    population = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

gdp_raw <- GET(gdp_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>% 
  pluck(2) %>% 
  as_tibble() %>% 
  select(
    country.value,
    year = date,
    gdp = value
  ) %>% 
  mutate(year = as.numeric(year)) %>% 
  filter(year %in% c(2019, 2020))

pop_and_gdp_raw <- inner_join(pop_raw, gdp_raw, by = c("country.value", "year")) %>% 
  mutate(country.value = recode(
           country.value,
          "Turkiye" = "Turkey",
          "Hong Kong SAR, China" = "Hong Kong",
          "Viet Nam" = "Vietnam",
          "Korea, Rep." = "South Korea",
          "United States" = "United States of America"
          ),
         country_code = normalize_name(country.value))

joined_all <- joined %>% 
  left_join(pop_and_gdp_raw, by = c("country_key" = "country_code", "year")) %>% 
  mutate(
    gdppercapita = gdp / population
  )
                              
```

# Main Dashboard


## RQ2 visuals

### visual 

```{r}
#| fig-alt: "A ridgeline density plot showing the distribution of plastic waste per capita in pieces per person by continent, with four ridges for Europe, Americas, Asia, and Africa. All distributions are right-skewed and peak near zero plastic items recorded per person. Europe has the widest spread with several secondary peaks and one extreme outlier. Asia also has an extreme outlier at 0.010 pieces of plastic per capita. Africa is the most concentrated near zero with a slightly wider base than the others."
#| fig-cap: "Density ridge plot showing the distribution of pieces of plastic collected per capita across regions. Each ridge represents the distribution of country-level plastic totals per capita within a region."

vis_1.5 <- joined_all %>%
  filter(!is.na(region), region != "Oceania") %>% 
  group_by(country, region, population) %>%
  summarize(
    total = sum(grand_total, na.rm = TRUE), .groups = "drop"
  ) %>% 
  mutate(
    region = fct_reorder(region, total, .desc = TRUE),
    waste_per_capita = total/population
  ) %>% 
  ggplot(aes(x=waste_per_capita, y=region, fill=region)) +
  geom_density_ridges(alpha = 0.5, height = 1, position = "identity") +
  labs(
    title = "Plastic Waste per Capita by Continent",
    subtitle = "Distribution of Plastic Waste per Capita by Continent (in Pieces/Person)",
    x= "Plastic Waste Per Capita (Items per Person)",
    y= ""
  ) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 15))

vis_1.5
```

### Tables {height="45%"}

::: panel-tabset
## Plastic Per Capita

</br>

Plastic Per Person

```{r}
#| label: top-30
joined_all %>%
  group_by(country, year) %>%
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    population = mean(population, na.rm = TRUE),
    plastic_per_person = total_plastic / population,
    .groups = "drop"
  ) %>%
  arrange(desc(plastic_per_person))%>%
  slice_head(n = 20) %>% 
  gt() %>% 
  cols_label(
    country = md("**Country**"),
    year = md("**Year**"),
    total_plastic = md("**Total Plastic**"),
    population = md("**Population**"),
    plastic_per_person = md("**Plastic per Person**")
  ) %>% 
  fmt_number(
    columns = c(total_plastic, population),
    decimals = 0
  ) %>% 
  fmt_number(
    columns = plastic_per_person,
    decimals = 5
  ) %>% 
  tab_header(
    title = md("**Top 20 Countries by Plastic Contributed per Person**")
  ) %>% 
  tab_source_note(
    source_note = "Plastic per person was calculated as total plastic collected divided by population."
  )
```

## GDP

</br>

GDP of Countries and GDP per capita

```{r}
#| label: bottom-30
gdp_combo <- joined_all %>%
  filter(!is.na(gdp), !is.na(gdppercapita), !is.na(country)) %>%
  group_by(country, year) %>%
  summarise(
    gdp = mean(gdp, na.rm = TRUE),
    gdp_per_capita = mean(gdppercapita, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(gdp))

gdp_combo %>%
  gt() %>%
  cols_label(
    country = md("**Country**"),
    year = md("**Year**"),
    gdp = md("**GDP (USD)**"),
    gdp_per_capita = md("**GDP per Capita (USD)**")
  ) %>%
  fmt_currency(
    columns = gdp,
    currency = "USD",
    decimals = 0
  ) %>%
  fmt_currency(
    columns = gdp_per_capita,
    currency = "USD",
    decimals = 0
  ) %>%
  tab_header(
    title = md("**GDP and GDP per Capita by Country and Year**")
  ) %>%
  tab_source_note(
    source_note = "GDP and GDP per capita values are reported in current U.S. dollars from the World Bank."
  )
```
:::

## Visual and model

### visual gdp {height="45%"}

```{r}
#| fig-alt: "A scatterplot comparing recorded plastic waste on the y-axis against GDP per capita on the x-axis, with points colored by continent. The majority of points are clustered near zero on both axes. One African country is a strong outlier with approximately 165,000 units of plastic waste recorded despite a very low GDP per capita below 5,000. Several Asian and Americas countries also show elevated waste counts at low GDP levels. European countries are spread across the full GDP per capita range but generally show low waste counts. No clear positive or negative relationship between GDP per capita and plastic waste is visible across the full dataset."
#| fig-cap: "Scatterplot showing the relationship between GDP per capita and total plastic collected across countries. Points are colored by region to highlight geographic differences."

vis_2 <- joined_all %>% 
  filter(!is.na(region), !is.na(gdppercapita)) %>%
  group_by(country, gdppercapita, region) %>% 
  summarize(
    total = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  ggplot(aes(x = gdppercapita, y = total, color = region)) +
  geom_point(alpha = 1, size = 1.5) +
  scale_x_continuous(labels = label_dollar()) +
  scale_color_brewer(
    palette = "Dark2",
    name = "World Region"
  ) +
  coord_cartesian(ylim = c(0, 180000)) +
  labs(
    title = "Higher-Income Countries Show Wide Variation in Plastic Waste",
    subtitle = "Country-level totals of recorded plastic waste vs GDP per capita",
    x = "GDP per Capita (USD)",
    y = "Total Plastic Waste"
  ) +
  theme_light(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size= 15),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

vis_2
```



### leaflet

```{r}
world_coords <- map_data("world") |>
  group_by(region) |>
  summarise(
    lng = mean(long, na.rm = TRUE),
    lat = mean(lat, na.rm = TRUE),
    .groups = "drop"
  )

plastics_country <- joined_all |>
  filter(parent_company == "Grand Total") |>
  group_by(country) |>
  summarise(
    total_plastic = sum(grand_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    country = recode(
      country,
      "Cote D_ivoire" = "Ivory Coast",
      "ECUADOR" = "Ecuador",
      "Hong Kong" = "Hong Kong",
      "NIGERIA" = "Nigeria",
      "United Kingdom" = "UK",
      "United States of America" = "USA"
    )
  ) |>
  left_join(world_coords, by = c("country" = "region"))

leaflet(plastics_country |> filter(!is.na(lat), !is.na(lng))) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~lng,
    lat = ~lat,
    radius = ~sqrt(total_plastic) / 25,
    popup = ~paste0(
      "<strong>", country, "</strong><br>",
      "Total Plastic: ", scales::comma(total_plastic)
    ),
    fillOpacity = 0.7,
    stroke = FALSE
  )
```


# Country Deep Dive

## New country

### Why this country?
We selected Nigeria because its plastic per capria waste was a very big outlier, making it useful for understanding general patterns in the dataset. How it differs from a typical country.


```{r}

renewable_code <- "EG.FEC.RNEW.ZS"  
pop_growth_code <- "SP.POP.GROW"
gdp_pc_code <- "NY.GDP.PCAP.CD"

renewable_url <- str_c(base_url, renewable_code, "?format=json&per_page=20000")
pop_growth_url <- str_c(base_url, pop_growth_code, "?format=json&per_page=20000")
gdp_pc_url <- str_c(base_url, gdp_pc_code, "?format=json&per_page=20000")

# Renewable energy %
renewable_raw <- GET(renewable_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  mutate(
    country = country.value,
    year = as.numeric(date),
    renewable_pct = value
  ) %>% 
  select(country, year, renewable_pct) %>% 
  filter(
    country == "Montenegro",
    year >= 2010,
    year <= 2020
  ) %>% 
  arrange(year)

#pop growth
pop_growth_raw <- GET(pop_growth_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>%
  pluck(2) %>% 
  as_tibble() %>% 
  mutate(
    country = country.value,
    year = as.numeric(date),
    pop_growth = value
  ) %>% 
  select(country, year, pop_growth) %>% 
  filter(
    country == "Montenegro",
    year >= 2010,
    year <= 2020
  ) %>% 
  arrange(year)

#gdp per capita
gdp_pc_raw <- GET(gdp_pc_url) %>% 
  content(as = "text") %>% 
  fromJSON(flatten = TRUE) %>% 
  pluck(2) %>% 
  as_tibble() %>% 
  mutate(
    country = country.value,
    year = as.numeric(date),
    gdp_pc = value
  ) %>% 
  select(country, year, gdp_pc) %>% 
  filter(
    country == "Montenegro",
    year >= 2010,
    year <= 2020
  ) %>% 
  arrange(year)

montenegro <- pop_growth_raw %>%
  left_join(gdp_pc_raw, by = c("country", "year"))

montenegro_full <- montenegro %>%
  left_join(renewable_raw, by = c("country", "year"))

```



## more

### Plot

```{r}
p <- ggplot(renewable_raw, aes(
  x = year,
  y = renewable_pct,
      group = 1,
  text = paste0("Year: ", year, "<br>Renewable: ", round(renewable_pct, 2), "%")
)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Renewable Energy Consumption in the Netherlands (2010-2020)",
    x = "Year",
    y = "Renewable Energy Consumption (%)"
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

### plot

```{r}

p_gdp <- ggplot(gdp_pc_raw, aes(
  x = year,
  y = gdp_pc,
  group = 1,
  text = paste0(
    "Year: ", year,
    "<br>GDP per Capita: $", round(gdp_pc, 0)
  )
)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "GDP per Capita in the Netherlands (2010-2020)",
    x = "Year",
    y = "GDP per Capita"
  ) +
  theme_minimal()

ggplotly(p_gdp, tooltip = "text")
```

### plot

```{r}
p_pop <- ggplot(pop_growth_raw, aes(
  x = year,
  y = pop_growth,
  group = 1,
  text = paste0(
    "Year: ", year,
    "<br>Population Growth: ", round(pop_growth, 2), "%"
  )
)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    title = "Population Growth in the Netherlands (2010-2020)",
    x = "Year",
    y = "Population Growth (%)"
  ) +
  theme_minimal()

ggplotly(p_pop, tooltip = "text")
```

